Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-21 ms/ms.Rnw
resubmitted

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 190\\
Introduction: 1179\\
Materials and Methods: 1176\\
Results: 433\\
Discussion: 1504\\
Acknowledgement: 24\\
3 Figures and 1 Table, 1 Supplemental Figure

@@

\linenumbers
\setstretch{\stretchy}
%\pagenumbering{gobble}

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution. First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange. These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
\end{itemize}

@@

Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r lifeform, stomata, stomatal ratio

@@

In answering these questions, I both reassessed previous hypotheses using newer phylogenetic comparative methods and evaluated previously untested hypotheses. I predicted \textit{a priori} that light habitat and growth would be correlated. Species with faster life histories, especially therophytes (annuals), would on average inhabitat sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees). Based on hypotheses from previous studies, I also predicted that herbaceous growth form and high light would be associated with amphistomy, even after controlling for phylogenetic nonindependence. Although these predictions have been tested previously, it is critical to reevaluate them here with updated methods because the subsequent untested hypotheses build on these results. The first novel hypothesis I tested predicts that light and growth form interact. Specifically, I hypothesized that both high light and herbaceous growth would be required to favor a more even stomatal ratio (i.e. amphistomy). Finally, I tested whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate. If high light and growth form favor amphistomy because it increases photosynthesis, then it follows that they should also favor other stomatal traits that reinforce this advantage. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would thus reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation. Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.

@@

I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms. I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. However, this analysis indicated no phylogenetic signal in the regression (See the R code accompanying this paper for further detail). Specifically, the estimated $\alpha$ parameter was extremely high. In the Ornstein-Uhlenbeck model, $\alpha$ is proportional to the inverse of the phylogenetic half-life (i.e. phylogenetic signal). When there is no phylogenetic signal (i.e. high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates. Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists performs worse than nonphylogenetic tests \citep{Revell_2010}. Therefore, I used a regular ANOVA with Type-2 sum of squares.

@@

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.

I also tested whether phylogenetic could explain the residual variation in stomatal ratio after accounting for growth form and \el. Specifically, I compared the expected residual variation given the actual tree versus a hypothetical tree where trait evolution has reached stationarity (i.e. a star phylogeny with infinite branch lengths). If phylogeny explains much of the variation, then the simulated residual variance from the actual tree should be greater than that of the stationary tree. I simulated trait values from 10$^4$ parametric bootstrap samples of the model with the lowest AIC (this was the model including Raunki\ae r lifeform, \el, and their interaction; see Results). I performed the first set of simulations using the actual phylogenetic tree in \pkg{OUwie} version 1.50 \citep{Beaulieu_OMeara_2016}. Each simulation used a different bootstrap parameter sample of $\alpha$ and $\sigma ^ 2$, where $\alpha$ is the return rate to the mean and $\sigma ^ 2$ is the diffusion rate. At stationarity, the variance of an Ornstein-Uhlenbeck trait is equal to $\sigma ^ 2 / 2 \alpha$. Therefore, I simulated stationary data by assuming a normal distribution with this variance estimated from the bootstrap samples. For comparability, I set the mean of simulations from both actual phylogeny and the stationary to zero. I compared the actual to stationary variance across simulated datasets using a paired $t$-test.

@@

I also tested whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface. The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$. The second scenario predicts that \el~acts similarly on both and that there is a negative covariance ($\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$). I tested these competing predictions by fitting a very simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}. In general, SEMs attempt to determine whether variables are related causally or whether a relationship is mediated by another correlated variable. Phylogenetic SEMs use the phylogenetic covariance matrix, as described above, rather than the raw covariance. Here, I used a phylogenetic SEM to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e. estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$). To fit the SEM, I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012}. I tested whether parameter estimates were significantly different from zero using $z$-scores.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el.

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary history. By bringing together datasets on stomata, light tolerance, growth form, and phylogeny of British angiosperms, I tested new hypotheses and reevaluated previous results using modern phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Fig.~\ref{fig:lf-light}, Fig.~\ref{fig:hf-light}). Nevertheless, both \el~and growth form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. Finally, I show for the first time that adaxial stomatal density in particular accounts for most of the coordinated evolution between light tolerance and stomatal density. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Among British angiosperms, phylogenetic comparative analyses suggest that selection favors amphistomy in high light habitats among fast-growing plants, but not shrubs and trees. This is a significant advance over previous studies that considered each factor in isolation and/or did not use modern approaches to control for phylogenetic nonindenpendence. For example, pioneering studies by \cite{Salisbury_1927} first suggested that amphistomy is associated with herbs in open habitats, albeit without formal statistical tools to disentangle these effects. Later work by \cite{Peat_Fitter_1994b} demonstrated these trends again using family-level comparisons, a basic method to account for phylogenetic nonindependence \citep[see also][]{Mott_etal_1982, Beerling_Kelly_1996}. However, this approach is still problematic because traits like growth from can be highly phylogenetically conserved. For example, orders like Fagales contain multiple families dominated by hypostomatous trees, hence it is premature to conclude that this correlation is biologically meaningful without properly accounting for phylogenetic nonidendence. By combining trait, ecological, and phylogenetic datasets on British angiosperms, we now know that not only do both light and growth form influence stomatal ratio, but in fact their effects may reinforce one another. Based on information criteria, light may be a more important factor than growth form or their interaction (Table~\ref{table:srAIC}), consistent with previous studies indicating a dominant role o flight \citep{Mott_etal_1982, Jordan_etal_2014, Bucher_etal_2017}.

The interaction between light and growth form among British angiosperms suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. Along these lines, Raunki\ae r lifeform may explain stomatal ratio better than plant habit (Table~\ref{table:srAIC}) because it is a better proxy for life history characteristics. For example, on an axis of `fast' to `slow' life history, geophytes more closely resemble phanerophytes than do chamaephytes or hemicryptophytes \citep{Salguero-Gomez_etal_2016}. Similarly, the relationship between light and stomatal ratio for geophytes was intermediate between that for phanerophytes and chamaephytes/hemicryptophytes (Fig.~\ref{fig:hf-light}). These comparisons indirectly suggest that both high light and fast life history are necessary to induce strong selection for amphistomy. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both therophyte/annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface. Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands \citep{Richardson_etal_2017}. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because greater leaf wetness and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (mostly tall trees) are almost always hypostomatous \citep[see also][]{Muir_2015}. Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSR_lf_a$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction. Despite there being phylogenetic signal, residual phylogenetic variation in stomatal ratio at the broad phylogenetic scale encompassed by British angiosperms should be at stationarity. The variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable at stationarity under an Ornstein-Uhlenbeck process (see Results). This may not be the case for younger clades that have radiated in the past few million years.

\subsection*{Why does adaxial stomatal density control stomatal ratio?}

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.

@@

\section*{Acknowledgements}
I thank Sally Otto, Matt Pennell, and Rob Salguero-G\'{o}mez for feedback on this manuscript. I was supported by an NSERC CREATE grant.

>>>>

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 199\\
Introduction: 1355\\
Materials and Methods: 1698\\
Results: 582\\
Discussion: 1935\\
Acknowledgement: 26\\
4 Figures and 1 Table, 5 Supplemental Figures

@@

\linenumbers
\setstretch{\stretchy}
%\pagenumbering{gobble}
%\raggedright

%\clearpage

@@

\begin{itemize}

	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.

	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio and density.

	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees). Furthermore, amphistomy and stomatal density evolve together in response to light.

	\item Comparative analyses of British angiosperms reveal two major insights. First, light and growth form interact to shape stomatal ratio; amphistomy is common under high light, but mostly for herbs. Second, coordinated evolution of adaxial stomatal density and light tolerance indicates that amphistomy helps to optimally balance light acquisition with gas exchange. Stomatal ratio may have potential as a functional trait for paleoecology and crop improvement.

\end{itemize}

@@

Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r life form, stomata, stomatal ratio

@@

In answering these questions, I both reassessed previous hypotheses using newer phylogenetic comparative methods and evaluated previously untested hypotheses. I predicted \textit{a priori} that light habitat and growth would be correlated. Species with faster life histories, especially therophytes (annuals), would on average inhabit sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees). Based on hypotheses from previous studies, I also predicted that herbaceous growth form and high light would be associated with amphistomy, even after controlling for phylogenetic nonindependence. Although these predictions have been tested previously, it is critical to reevaluate them here with updated methods because the subsequent untested hypotheses build on these results. The first novel hypothesis I tested predicts that light and growth form interact. Specifically, I hypothesized that both high light and herbaceous growth would be required to favor a more even stomatal ratio (i.e. amphistomy). Finally, I tested whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate. If high light and growth form favor amphistomy because it increases photosynthesis, then it follows that they should also favor other stomatal traits that reinforce this advantage. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would thus reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation. Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.

@@

I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms. I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. However, this analysis indicated no phylogenetic signal in the regression (See the R code accompanying this paper for further detail). Specifically, the estimated $\alpha$ parameter was extremely high. In the Ornstein-Uhlenbeck model, $\alpha$ is proportional to the inverse of the phylogenetic half-life (i.e. phylogenetic signal). When there is no phylogenetic signal (i.e. high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates. Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists perform worse than nonphylogenetic tests \citep{Revell_2010}. Therefore, I used a regular ANOVA with Type-2 sum of squares.

@@

% REMOVED: I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface.

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC), a common measure of model fit that penalizes additional parameters. Phylogenetic linear models simultaneously estimate the effect of continuous and categorical predictors while controlling for phylogenetic nonindependence. For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. The Ornstein-Uhlenbeck model is characterized by a diffusion rate ($\sigma ^ 2$) and a return rate ($\alpha$), which describes the phylogenetic signal (see above). I used 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.

I tested whether phylogenetic nonstationarity could explain the residual variation in stomatal ratio after accounting for growth form and \el. Specifically, I compared the expected residual variation given the actual tree versus a hypothetical tree where trait evolution has reached stationarity (i.e. a star phylogeny with infinite branch lengths). If phylogeny explains much of the variation, then the simulated residual variance from the actual tree should be greater than that of the stationary tree. I simulated trait values from 10$^4$ parametric bootstrap samples of the model with the lowest AIC (this was the model including Raunki\ae r life form, \el, and their interaction; see Results). I performed the first set of simulations using the actual phylogenetic tree in \pkg{OUwie} version 1.50 \citep{Beaulieu_OMeara_2016}. Each simulation used a different bootstrap parameter sample of $\alpha$ and $\sigma ^ 2$, where $\alpha$ is the return rate to the mean and $\sigma ^ 2$ is the diffusion rate. At stationarity, the variance of an Ornstein-Uhlenbeck trait is equal to $\sigma ^ 2 / 2 \alpha$. Therefore, I simulated stationary data by assuming a normal distribution with this variance estimated from the bootstrap samples. For comparability, I set the mean of simulations from both actual phylogeny and the stationary `phylogeny' to zero. I compared the actual to stationary variance across simulated datasets using a paired $t$-test.

@@

If light-mediated increases in adaxial stomatal density can evolve while abaxial density remains roughly constant, then the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ will be stronger than that for $\mathrm{sd_{ab}}$. Under this scenario, stomatal ratio and density evolve in a coordinated fashion in response to light. Alternatively, if greater \el~favors greater stomatal ratio but total stomatal density is roughly constant, then there will be a negative covariance between ab- and adaxial density ($\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$). I tested these competing predictions by fitting a simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}. In general, SEMs attempt to determine whether variables are related causally or whether a relationship is mediated by another correlated variable. Phylogenetic SEMs use the phylogenetic covariance matrix, as described above, rather than the raw covariance. Here, I used a phylogenetic SEM to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e. estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$). I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012} to fit the SEM by finding parameter estimates would lead to phylogenetic covariance close to that observed in the data. I tested whether parameter estimates were significantly different from zero using $z$-scores.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el~less strongly ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore I focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \els. In contrast, phanerophytes were nearly always hypostomatous regardless of \el.

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance of this variation is not well understood. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary history. By bringing together datasets on stomata, light tolerance, growth form, and phylogeny of British angiosperms, I tested new hypotheses and reevaluated previous results using modern phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Fig.~\ref{fig:lf-light}, Fig.~\ref{fig:hf-light}). Nevertheless, both \el~and growth form affect stomatal ratio, but these factors also interact. This new finding shows that the influence of \el~on stomatal ratio varies across forms. Finally, I show for the first time that adaxial stomatal density in particular accounts for most of the coordinated evolution between light tolerance and stomatal density. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Among British angiosperms, phylogenetic comparative analyses suggest that selection favors amphistomy in high light habitats among fast-growing herbs, but not shrubs and trees. This is a significant advance over previous studies that considered each factor in isolation and/or did not use modern approaches to control for phylogenetic nonindependence. For example, pioneering studies by \cite{Salisbury_1927} first suggested that amphistomy is associated with herbs in open habitats, albeit without formal statistical tools to disentangle light and growth form. Later work by \cite{Peat_Fitter_1994b} demonstrated these trends again using family-level comparisons, a basic method to account for phylogenetic nonindependence \citep[see also][]{Mott_etal_1982, Beerling_Kelly_1996}. However, this approach is still problematic because traits like growth from can be highly phylogenetically conserved. For example, orders like Fagales contain multiple families dominated by hypostomatous trees, hence it is premature to conclude that this correlation is biologically meaningful without properly accounting for phylogenetic nonindependence. By combining trait, ecological, and phylogenetic datasets on British angiosperms, we now know that not only do both light and growth form influence stomatal ratio, but in fact their effects may reinforce one another. Based on information criteria, light may be a more important factor than growth form or their interaction (Table~\ref{table:srAIC}), consistent with previous studies indicating a dominant role of light \citep{Mott_etal_1982, Jordan_etal_2014, Bucher_etal_2017}.

The interaction between light and growth form among British angiosperms suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. Along these lines, Raunki\ae r life form may explain stomatal ratio better than plant habit (Table~\ref{table:srAIC}) because it is a better proxy for life history characteristics. For example, on an axis of `fast' to `slow' life history, geophytes more closely resemble phanerophytes than do chamaephytes or hemicryptophytes \citep{Salguero-Gomez_etal_2016}. Similarly, the relationship between light and stomatal ratio for geophytes was intermediate between that for phanerophytes and chamaephytes/hemicryptophytes (Fig.~\ref{fig:hf-light}). These comparisons indirectly suggest that both high light and fast life history are necessary to induce strong selection for amphistomy. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both therophyte/annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of palisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface. Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands \citep{Richardson_etal_2017}. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because wetter leaves and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes are almost always hypostomatous \citep[see also][]{Muir_2015}. Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSR_lf_a$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction. Despite there being phylogenetic signal, residual phylogenetic variation in stomatal ratio at the broad phylogenetic scale encompassed by British angiosperms should be at stationarity. The observed variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable from that expected for a trait at stationarity under an Ornstein-Uhlenbeck process (see Results). This may not be the case for younger clades that have radiated in the past few million years.

\subsection*{Coordinated evolution of stomatal ratio and density in response to light}

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. To my knowledge, this highly nonrandom pattern among British angiosperms has not been demonstrated before, but it parallels evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.

@@

\section*{Acknowledgements}
I thank Sally Otto, Matt Pennell, Rob Salguero-G\'{o}mez, and two anonymous reviewers for feedback on this manuscript. I was supported by an NSERC CREATE grant.
